A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-09-07 paper/paper.intro.tex
Fix most chktex warnings

Automated methods to optimize schedules have been studied and applied in many
industries, including transportation \cite{aickelin_improved_2006,
	goel_truck_2012, gunther_combined_2010}, manufacturing
\cite{al-yakoob_mixed-integer_2007, al-yakoob_column_2008,
	alfares_simulation_2007}, retail \cite{chapados_retail_2011,
	nissen_automatic_2010} and military \cite{horn_scheduling_2007,
	laguna_modeling_2005}. Of special interest to a clinician on-call scheduling
problem are the approaches to schedule nurses, who often work in shifts. In the
nurse scheduling problem, the goal is to find an optimal assignment of nurses to
shifts that satisfies all of the hard constraints, such as hospital regulations,
and as many soft constraints as possible, which may include nurse preferences. A
wide variety of approaches, including exact and heuristic approaches, have been
used to solve the nurse scheduling problem: integer linear programming
\cite{azaiez_0-1_2005, trilling_nurse_2006, widyastiti_nurses_2016}, network
flows \cite{el_adoly_new_2018}, genetic algorithms
\cite{aickelin_exploiting_2000, jan_evolutionary_2000, kawanaka_genetic_2001},
simulated annealing \cite{jaszkiewicz_metaheuristic_1997}, and artificial
intelligence \cite{abdennadher_nurse_nodate, li_hybrid_2003}. A comprehensive
literature review of these and other methods applied to nurse rostering is
presented in \cite{burke_state_2004}. \\

@@

Many of the approaches to nurse scheduling were designed to satisfy the
requirements of a specific hospital department which causes a large number of
variables and constraints to be incorporated into the problem formulation. While
these department-specific approaches allow end-users to find precise schedules
that satisfy the needs of the department and the preferences of the nurses and
clinicians in that department, they are difficult to readily adapt to other
departments in the same hospital or other hospitals. % explain why hard to
%adapt?... what makes their generalizablility/adaptability limited?
Moreover, the large number of variables and constraints also leads to
computational complexity issues \cite{goos_complexity_1996}, especially when
using exact methods for finding the solution. In this paper, we tackle a version
of the nurse scheduling problem arising from a case study of one clinical
division, providing two different services simultaneously (general infectious
disease (ID) consults; and HIV consults service) at St. Michael's Hospital in
Toronto, Canada. Our goal is to (1) present a simple integer linear programming
(ILP) formulation for the scheduling problem as developed for the hospital, and
describe the adaptability of the formulation to solving similar problems in
other departments; (2) compare the performance of the ILP scheduler to the
results of the manual approach; and (3) analyze the robustness of the algorithm
in difficult instances of the problem. \\
% present a simple formulation for the problem developed and tested at the
%hospital after switching from a manual approach to scheduling; and (2) analyze
%the performance of integer linear programming in solving difficult instances of
%the problem and compare the results with those of the manual approach; and (3)
%describe the adaptability of the formulation as a basic framework for solving
%similar problems in other departments. \\ %make #3 an objective

We begin by describing the problem in Section \ref{sec:problem}, and presenting
our ILP formulation in Section \ref{sec:methods}. Next, we compare the results
of our formulation to manually-created schedules, and evaluate the performance
of the algorithm on simulated data in Section \ref{sec:results}. Finally, we
discuss and interpret the results in Section \ref{sec:discussion}. % list the
%main contents/elements of the paper.

>>>>

Automated methods to optimize schedules have been studied and applied in many
industries, including transportation~\cite{aickelin_improved_2006,
	goel_truck_2012, gunther_combined_2010}, manufacturing~\cite{al-yakoob_mixed-integer_2007, al-yakoob_column_2008,
	alfares_simulation_2007}, retail~\cite{chapados_retail_2011,
	nissen_automatic_2010} and military~\cite{horn_scheduling_2007,
	laguna_modeling_2005}. Of special interest to a clinician on-call scheduling
problem are the approaches to schedule nurses, who often work in shifts. In the
nurse scheduling problem, the goal is to find an optimal assignment of nurses to
shifts that satisfies all of the hard constraints, such as hospital regulations,
and as many soft constraints as possible, which may include nurse preferences. A
wide variety of approaches, including exact and heuristic approaches, have been
used to solve the nurse scheduling problem: integer linear programming~\cite{azaiez_0-1_2005, trilling_nurse_2006, widyastiti_nurses_2016}, network
flows~\cite{el_adoly_new_2018}, genetic algorithms~\cite{aickelin_exploiting_2000, jan_evolutionary_2000, kawanaka_genetic_2001},
simulated annealing~\cite{jaszkiewicz_metaheuristic_1997}, and artificial
intelligence~\cite{abdennadher_nurse_nodate, li_hybrid_2003}. A comprehensive
literature review of these and other methods applied to nurse rostering is
presented in~\cite{burke_state_2004}. \\

@@

Many of the approaches to nurse scheduling were designed to satisfy the
requirements of a specific hospital department which causes a large number of
variables and constraints to be incorporated into the problem formulation. While
these department-specific approaches allow end-users to find precise schedules
that satisfy the needs of the department and the preferences of the nurses and
clinicians in that department, they are difficult to readily adapt to other
departments in the same hospital or other hospitals. % explain why hard to
%adapt?... what makes their generalizablility/adaptability limited?
Moreover, the large number of variables and constraints also leads to
computational complexity issues~\cite{goos_complexity_1996}, especially when
using exact methods for finding the solution. In this paper, we tackle a version
of the nurse scheduling problem arising from a case study of one clinical
division, providing two different services simultaneously (general infectious
disease (ID) consults; and HIV consults service) at St.\ Michael's Hospital in
Toronto, Canada. Our goal is to (1) present a simple integer linear programming
(ILP) formulation for the scheduling problem as developed for the hospital, and
describe the adaptability of the formulation to solving similar problems in
other departments; (2) compare the performance of the ILP scheduler to the
results of the manual approach; and (3) analyze the robustness of the algorithm
in difficult instances of the problem. \\
% present a simple formulation for the problem developed and tested at the
%hospital after switching from a manual approach to scheduling; and (2) analyze
%the performance of integer linear programming in solving difficult instances of
%the problem and compare the results with those of the manual approach; and (3)
%describe the adaptability of the formulation as a basic framework for solving
%similar problems in other departments. \\ %make #3 an objective

We begin by describing the problem in Section~\ref{sec:problem}, and presenting
our ILP formulation in Section~\ref{sec:methods}. Next, we compare the results
of our formulation to manually-created schedules, and evaluate the performance
of the algorithm on simulated data in Section~\ref{sec:results}. Finally, we
discuss and interpret the results in Section~\ref{sec:discussion}. % list the
%main contents/elements of the paper.

### 2019-09-07 paper/paper.methods.tex
Fix most chktex warnings

In this section, we present an application of ILP to solve the clinician
scheduling problem presented in Section \ref{sec:problem}. First, we describe
the sets, indices and variables present in the formulation of the program. We
then convert the constraints given in Table \ref{tbl:constraint-summary} into
mathematical terms, and outline the objective function of the linear program.

\subsection{Sets and Indices} \label{sec:meth-sets-indices}
We denote the set of all services %or I think services is good (vs.
%services/divisions) as you define two types of services (ID vs. HIV) in the
%problem statement
that clinicians in a single department can provide as $\mathcal{S}$. The
following formulation assumes that all clinicians are able to provide all
services. The set of all clinicians in the department is denoted as
$\mathcal{C}$. The sets of blocks and weekends that clinicians will be assigned
to are denoted as $\mathcal{B}$ and $\mathcal{W}$ respectively. The block size
used in our experiments is 2 weeks, but the following LP formulation does not
specify a particular size for blocks, and so it can be adapted to the needs of
the department in question.
%The size of a block is not constrained and can be adapted %how?
%to the needs of the given department. 
A subset of weekends are denoted $\mathcal{L}$, corresponding to established
long/holiday weekends such as the Canadian Civic Day weekend in August. Lastly,
we denote with $\mathcal{BR}_c$ and $\mathcal{WR}_c$ the block and weekend
requests of clinicians as subsets of all blocks and weekends, respectively. For
instance, if clinician $c$'s requests intersect with blocks 1 and 2, and weekend
1, then $\mathcal{BR}_c = \{1, 2\}$ and $\mathcal{WR}_c = \{1\}$. Table
\ref{tbl:sets-indices} presents a summary of the sets and indices described.

\begin{table}[h]
	\centering
	\begin{tabular}{ c c l }
		\hline
		\textbf{Set}                         & \textbf{Index} & \textbf{Description}  
		\\ \hline
		$\mathcal{S} = \{1, \ldots, S \}$    & $s$            & services              
		\\
		$\mathcal{C} = \{1, \ldots, C \}$    & $c$            & clinicians            
		\\
		$\mathcal{B} = \{1, \ldots, B \}$    & $b$            & blocks                
		\\
		$\mathcal{W} = \{1, \ldots, W \}$    & $w$            & weekends              
		\\
		$\mathcal{L} \subset \mathcal{W}$    &                & long weekends         
		\\
		$\mathcal{BR}_c \subset \mathcal{B}$ &                & block requests of
		clinician $c$   \\
		$\mathcal{WR}_c \subset \mathcal{W}$ &                & weekend requests of
		clinician $c$
	\end{tabular}
	\caption{Description of sets and indices in the problem}
	\label{tbl:sets-indices}
\end{table}

\subsection{Variables} \label{sec:meth-variables}
Since each clinician may be assigned to work on any service, during any block of
the year, we denote such an assignment as a binary variable $X_{c, b, s}$. A
value of 1 indicates that the given clinician $c$ is assigned to provide service
$s$ during block $b$. Weekend assignments are similarly defined using a binary
variable $Y_{c, w}$, but without a service index, as clinicians are expected to
provide all services during the weekends. We then introduce a set of constants
$m_{c, s}$ and $M_{c, s}$ to constrain the minimal and maximal number of blocks
each clinician is allowed to work during the year. Table
\ref{tbl:variables-constants} presents a summary of the constants and variables
in the problem.
%To optimize the soft constraint Block-Weekend Adjacency, we maximize the
%product $X_{c, b, s} \cdot Y_{c, w}$ for adjacent blocks and weekends. To
%formulate such an objective as a linear function of variables, we introduce
%another set of variables, denoted by $Z_{c, b, s}$, with additional constraints
%on its range. Further details regarding this variable are described in Section
%\ref{sec:meth-objectives}.

\begin{table}[h]
	\centering
	\begin{tabular}{ c l }
		\hline
		\textbf{Name}              & \textbf{Description}                             
		\\ \hline
		$X_{c, b, s} \in \{0, 1\}$ & assignment of clinician $c$ to service $s$ on
		block $b$            \\
		$Y_{c, w} \in \{0, 1\}$    & assignment of clinician $c$ on weekend $w$       
		\\
		%		$Z_{c, b, s} \in \{0, 1\}$ & helper variable for optimizing Block-Weekend
		%adjacency             \\
		$m_{c, s}$                 & minimum number of blocks clinician $c$ should
		cover on service $s$ \\
		$M_{c, s}$                 & maximum number of blocks clinician $c$ should
		cover on service $s$
	\end{tabular}
	\caption{Description of variables and constants in the problem}
	\label{tbl:variables-constants}
\end{table}

\subsection{Constraints} \label{sec:meth-constraints}
We now formalize the hard constraints in Table \ref{tbl:constraint-summary}
using the variables defined above.

@@

\subsection{Objectives} \label{sec:meth-objectives}
As described in Section \ref{sec:problem}, the soft constraints of the clinician
scheduling problem include: satisfying clinician block off requests (BR),
satisfying clinician weekend off requests (WR), and assigning weekends closer to
blocks (BWA). We convert these soft constraints into linear objective functions
of the binary variables defined in Section \ref{sec:meth-variables}. Objectives
(\ref{eqn:obj-block-requests}) and (\ref{eqn:obj-weekend-requests}) can be
written as the following linear functions:  %minor suggestion - consider
%numbering the objectives for ease later when saying things like "these two
%objectives"...or writing the full name of the objectives instead of 'this' and
%'these' as sometimes can get confusing what the 'this' and the 'these' refer
%to. 
\begin{align}
&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S}
(-1)^{\ind(b\,\in\,\mathcal{BR}_c)}\cdot X_{c, b, s} \tag{BR}
\label{eqn:obj-block-requests}\\
&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W}
(-1)^{\ind(w\,\in\,\mathcal{WR}_c)}\cdot Y_{c, w} \tag{WR}
\label{eqn:obj-weekend-requests}
\end{align}
where $\ind(P)$ is the indicator function that has value 1 when predicate $P$
holds and 0 otherwise. In the above two objectives, we penalize any assignments
that conflict with a block or weekend request, and aim to maximize the
non-conflicting assignments. \\

@@

However, as it is, $Q_3$ is not a linear function of its variables and cannot be
optimized in a linear programming framework. An approach used to convert such
functions into linear objectives involves introducing a helper variable and
additional constraints \cite{hammer_boolean_1968}. In our case, introducing a
variable $Z_{c, b, s}$ for every product $X_{c, b, s} \cdot Y_{c, w}$ with $w =
\varphi(b)$, and constraining $Z$ such that 
\begin{align}
&Z_{c, b, s} \leq X_{c, b, s} \label{eqn:helper-x-constraint}\\
&Z_{c, b, s} \leq Y_{c, w=\varphi(b)} &&\forall s\in\mathcal{S}
\label{eqn:helper-y-constraint}
\end{align}
allows us to rewrite $Q_3$ as a linear function of $Z$,
\begin{align}
&Q_3(Z) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S} Z_{c, b, s}
\end{align}
Indeed from Eqns. (\ref{eqn:helper-x-constraint}) and
(\ref{eqn:helper-y-constraint}),  whenever $X_{c, b, s} \cdot Y_{c, w} = 1$
(respectively, 0), $Z_{c, b, s}$ can attain a maximum value of 1 (respectively,
0), giving us the correct adjacency maximization objective. \\
%Indeed, whenever $X_{c, b, s} \cdot Y_{c, w} = 1$, $Z_{c, b, s}$ can attain a
%maximum value of 1, and whenever $X_{c, b, s} \cdot Y_{c, w} = 0$, at least one
%of $X_{c, b, s}$ or $Y_{c, w}$ must be 0, so $Z_{c, b, s}$ will be constrained
%to attain a maximum value of 0, giving us the correct adjacency maximization
%objective. \\

In order to optimize all objectives simultaneously, we optimize a weighted sum
of the normalized objective functions,
\begin{equation}
\max_{X, Y, Z} \alpha \bar{Q}_1(X) + \beta \bar{Q}_2(Y) + (1 - \alpha - \beta)
\bar{Q}_3(Z)
\end{equation}
where $\bar{Q}_i$ is the normalization of objective $Q_i$, and $0 \leq \alpha,
\beta \leq 1$. This method guarantees an optimal solution to be Pareto optimal
\cite{stanimirovic_linear_2011}. \\

>>>>

In this section, we present an application of ILP to solve the clinician
scheduling problem presented in Section~\ref{sec:problem}. First, we describe
the sets, indices and variables present in the formulation of the program. We
then convert the constraints given in Table~\ref{tbl:constraint-summary} into
mathematical terms, and outline the objective function of the linear program.

\subsection{Sets and Indices}\label{sec:meth-sets-indices}
We denote the set of all services %or I think services is good (vs.
%services/divisions) as you define two types of services (ID vs. HIV) in the
%problem statement
that clinicians in a single department can provide as $\mathcal{S}$. The
following formulation assumes that all clinicians are able to provide all
services. The set of all clinicians in the department is denoted as
$\mathcal{C}$. The sets of blocks and weekends that clinicians will be assigned
to are denoted as $\mathcal{B}$ and $\mathcal{W}$ respectively. The block size
used in our experiments is 2 weeks, but the following LP formulation does not
specify a particular size for blocks, and so it can be adapted to the needs of
the department in question.
%The size of a block is not constrained and can be adapted %how?
%to the needs of the given department. 
A subset of weekends are denoted $\mathcal{L}$, corresponding to established
long/holiday weekends such as the Canadian Civic Day weekend in August. Lastly,
we denote with $\mathcal{BR}_c$ and $\mathcal{WR}_c$ the block and weekend
requests of clinicians as subsets of all blocks and weekends, respectively. For
instance, if clinician $c$'s requests intersect with blocks 1 and 2, and weekend
1, then $\mathcal{BR}_c = \{1, 2\}$ and $\mathcal{WR}_c = \{1\}$. Table~\ref{tbl:sets-indices} presents a summary of the sets and indices described.

\begin{table}[h]
	\centering
	\begin{tabular}{ c c l }
		\hline
		\textbf{Set}                         & \textbf{Index} & \textbf{Description}  
		\\ \hline
		$\mathcal{S} = \{1, \ldots, S \}$    & $s$            & services              
		\\
		$\mathcal{C} = \{1, \ldots, C \}$    & $c$            & clinicians            
		\\
		$\mathcal{B} = \{1, \ldots, B \}$    & $b$            & blocks                
		\\
		$\mathcal{W} = \{1, \ldots, W \}$    & $w$            & weekends              
		\\
		$\mathcal{L} \subset \mathcal{W}$    &                & long weekends         
		\\
		$\mathcal{BR}_c \subset \mathcal{B}$ &                & block requests of
		clinician $c$   \\
		$\mathcal{WR}_c \subset \mathcal{W}$ &                & weekend requests of
		clinician $c$
	\end{tabular}
	\caption{Description of sets and indices in the problem}\label{tbl:sets-indices}
\end{table}

\subsection{Variables}\label{sec:meth-variables}
Since each clinician may be assigned to work on any service, during any block of
the year, we denote such an assignment as a binary variable $X_{c, b, s}$. A
value of 1 indicates that the given clinician $c$ is assigned to provide service
$s$ during block $b$. Weekend assignments are similarly defined using a binary
variable $Y_{c, w}$, but without a service index, as clinicians are expected to
provide all services during the weekends. We then introduce a set of constants
$m_{c, s}$ and $M_{c, s}$ to constrain the minimal and maximal number of blocks
each clinician is allowed to work during the year. Table~\ref{tbl:variables-constants} presents a summary of the constants and variables
in the problem.
%To optimize the soft constraint Block-Weekend Adjacency, we maximize the
%product $X_{c, b, s} \cdot Y_{c, w}$ for adjacent blocks and weekends. To
%formulate such an objective as a linear function of variables, we introduce
%another set of variables, denoted by $Z_{c, b, s}$, with additional constraints
%on its range. Further details regarding this variable are described in Section
%\ref{sec:meth-objectives}.

\begin{table}[h]
	\centering
	\begin{tabular}{ c l }
		\hline
		\textbf{Name}              & \textbf{Description}                             
		\\ \hline
		$X_{c, b, s} \in \{0, 1\}$ & assignment of clinician $c$ to service $s$ on
		block $b$            \\
		$Y_{c, w} \in \{0, 1\}$    & assignment of clinician $c$ on weekend $w$       
		\\
		%		$Z_{c, b, s} \in \{0, 1\}$ & helper variable for optimizing Block-Weekend
		%adjacency             \\
		$m_{c, s}$                 & minimum number of blocks clinician $c$ should
		cover on service $s$ \\
		$M_{c, s}$                 & maximum number of blocks clinician $c$ should
		cover on service $s$
	\end{tabular}
	\caption{Description of variables and constants in the problem}\label{tbl:variables-constants}
\end{table}

\subsection{Constraints}\label{sec:meth-constraints}
We now formalize the hard constraints in Table~\ref{tbl:constraint-summary}
using the variables defined above.

@@

\subsection{Objectives}\label{sec:meth-objectives}
As described in Section~\ref{sec:problem}, the soft constraints of the clinician
scheduling problem include: satisfying clinician block off requests (BR),
satisfying clinician weekend off requests (WR), and assigning weekends closer to
blocks (BWA). We convert these soft constraints into linear objective functions
of the binary variables defined in Section~\ref{sec:meth-variables}. Objectives
(\ref{eqn:obj-block-requests}) and (\ref{eqn:obj-weekend-requests}) can be
written as the following linear functions:  %minor suggestion - consider
%numbering the objectives for ease later when saying things like "these two
%objectives"...or writing the full name of the objectives instead of 'this' and
%'these' as sometimes can get confusing what the 'this' and the 'these' refer
%to. 
\begin{align}
&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S}
{(-1)}^{\ind(b\,\in\,\mathcal{BR}_c)}\cdot X_{c, b, s} \tag{BR}
\label{eqn:obj-block-requests}\\
&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W}
{(-1)}^{\ind(w\,\in\,\mathcal{WR}_c)}\cdot Y_{c, w} \tag{WR}
\label{eqn:obj-weekend-requests}
\end{align}
where $\ind(P)$ is the indicator function that has value 1 when predicate $P$
holds and 0 otherwise. In the above two objectives, we penalize any assignments
that conflict with a block or weekend request, and aim to maximize the
non-conflicting assignments. \\

@@

However, as it is, $Q_3$ is not a linear function of its variables and cannot be
optimized in a linear programming framework. An approach used to convert such
functions into linear objectives involves introducing a helper variable and
additional constraints~\cite{hammer_boolean_1968}. In our case, introducing a
variable $Z_{c, b, s}$ for every product $X_{c, b, s} \cdot Y_{c, w}$ with $w =
\varphi(b)$, and constraining $Z$ such that 
\begin{align}
&Z_{c, b, s} \leq X_{c, b, s} \label{eqn:helper-x-constraint}\\
&Z_{c, b, s} \leq Y_{c, w=\varphi(b)} &&\forall s\in\mathcal{S}
\label{eqn:helper-y-constraint}
\end{align}
allows us to rewrite $Q_3$ as a linear function of $Z$,
\begin{align}
&Q_3(Z) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S} Z_{c, b, s}
\end{align}
Indeed from Eqns. (\ref{eqn:helper-x-constraint}) and
(\ref{eqn:helper-y-constraint}),  whenever $X_{c, b, s} \cdot Y_{c, w} = 1$
(respectively, 0), $Z_{c, b, s}$ can attain a maximum value of 1 (respectively,
0), giving us the correct adjacency maximization objective. \\
%Indeed, whenever $X_{c, b, s} \cdot Y_{c, w} = 1$, $Z_{c, b, s}$ can attain a
%maximum value of 1, and whenever $X_{c, b, s} \cdot Y_{c, w} = 0$, at least one
%of $X_{c, b, s}$ or $Y_{c, w}$ must be 0, so $Z_{c, b, s}$ will be constrained
%to attain a maximum value of 0, giving us the correct adjacency maximization
%objective. \\

In order to optimize all objectives simultaneously, we optimize a weighted sum
of the normalized objective functions,
\begin{equation}
\max_{X, Y, Z} \alpha \bar{Q}_1(X) + \beta \bar{Q}_2(Y) + (1 - \alpha - \beta)
\bar{Q}_3(Z)
\end{equation}
where $\bar{Q}_i$ is the normalization of objective $Q_i$, and $0 \leq \alpha,
\beta \leq 1$. This method guarantees an optimal solution to be Pareto optimal~\cite{stanimirovic_linear_2011}. \\

### 2019-09-07 paper/paper.problem.tex
Fix most chktex warnings

At St. Michael's Hospital, the division of infectious diseases (ID) offers
general ID and HIV consultation on inpatients as two parallel services. Each
service provides clinical care throughout the year, during regular work weeks as
well as weekends and holidays. The clinicians in the division typically receive
a schedule in advance, outlining their on-call service dates for the full year.
In the yearly schedule each clinician is assigned to blocks of regular work
weeks and weekends. Each block corresponds to two consecutive work weeks. Apart
from long (holiday) weekends, a work week starts on Monday at 8 A.M. and ends on
Friday at 5 P.M. Accordingly, weekend service starts on Friday at 5 P.M., and
ends on Monday at 8 A.M. During the weekend, ID and HIV consultation services
are combined and provided by one clinician. During the regular work week the ID
and HIV services are led by one clinician each. \\

@@

In addition to maintaining a balanced work load among clinicians, the schedule
should also accommodate their preferences. Clinicians provide their requests for
time off ahead of schedule generation so that the requests may be integrated
into the schedule. Individuals may specify days, weeks or weekends off, with the
understanding that any blocks overlapping with their request will be assigned to
a different clinician where possible. For example, if a clinician only requests
a given Monday and Tuesday off, the schedule will generally avoid assigning the
entire block to that clinician. Clinicians also prefer to have their weekend and
block assignments close together, so the schedule should account for this when
distributing assignments. A summary of the outlined constraints is given in
Table \ref{tbl:constraint-summary}.

\begin{table}[h]
	\centering
	\begin{tabular}{ l l l }
		\hline
		\textbf{Constraint Name} & \textbf{Description}                               
		&
		\textbf{Type} \\ \hline
		Block Coverage (BC)           & \makecell[l]{each service needs to have
			exactly \\ one clinician that covers any given block}                           
		& Hard          \\ \hline
		Weekend Coverage (WC)        & \makecell[l]{every weekend needs to have
			exactly \\ one clinician that covers it}                                        
		& Hard          \\ \hline
		Min/Max (MM)                  & \makecell[l]{for a given service, each
			clinician can only \\ work between the minimum and maximum \\ number of allowed
			blocks} & Hard          \\ \hline
		No Consecutive Blocks (NCB)   & \makecell[l]{any clinician should not work \\
			two consecutive blocks, across all services}                                   
		& Hard          \\ \hline
		No Consecutive Weekends (NCW) & \makecell[l]{any clinician should not work two
			consecutive \\ weekends}                                                       
		& Hard          \\ \hline
		Equal Weekends (EW)          & \makecell[l]{weekends should be equally
			distributed \\ between clinicians}                                              
		& Hard          \\ \hline
		Equal Holidays (EH)          & \makecell[l]{long weekends should be equally
			distributed \\ between clinicians}                                              
		& Hard          \\ \hline
		Block Requests (BR)         & \makecell[l]{each clinician can request to be
			off service \\ during certain blocks throughout the year}                       
		& Soft          \\ \hline
		Weekend Requests (WR)        & \makecell[l]{each clinician can request to be
			off service \\ during certain weekends throughout the year}                     
		& Soft          \\ \hline
		Block-Weekend Adjacency (BWA)  & \makecell[l]{the block and weekend
			assignments of a given \\ clinician should be adjacent}                         
		& Soft
	\end{tabular}
	\caption{Summary of the constraints for the clinician scheduling problem}
	\label{tbl:constraint-summary}
\end{table}

>>>>

At St.\ Michael's Hospital, the division of infectious diseases (ID) offers
general ID and HIV consultation on inpatients as two parallel services. Each
service provides clinical care throughout the year, during regular work weeks as
well as weekends and holidays. The clinicians in the division typically receive
a schedule in advance, outlining their on-call service dates for the full year.
In the yearly schedule each clinician is assigned to blocks of regular work
weeks and weekends. Each block corresponds to two consecutive work weeks. Apart
from long (holiday) weekends, a work week starts on Monday at 8 A.M. and ends on
Friday at 5 P.M. Accordingly, weekend service starts on Friday at 5 P.M., and
ends on Monday at 8 A.M. During the weekend, ID and HIV consultation services
are combined and provided by one clinician. During the regular work week the ID
and HIV services are led by one clinician each. \\

@@

In addition to maintaining a balanced work load among clinicians, the schedule
should also accommodate their preferences. Clinicians provide their requests for
time off ahead of schedule generation so that the requests may be integrated
into the schedule. Individuals may specify days, weeks or weekends off, with the
understanding that any blocks overlapping with their request will be assigned to
a different clinician where possible. For example, if a clinician only requests
a given Monday and Tuesday off, the schedule will generally avoid assigning the
entire block to that clinician. Clinicians also prefer to have their weekend and
block assignments close together, so the schedule should account for this when
distributing assignments. A summary of the outlined constraints is given in
Table~\ref{tbl:constraint-summary}.

\begin{table}[h]
	\centering
	\begin{tabular}{ l l l }
		\hline
		\textbf{Constraint Name} & \textbf{Description}                               
		&
		\textbf{Type} \\ \hline
		Block Coverage (BC)           & \makecell[l]{each service needs to have
			exactly \\ one clinician that covers any given block}                           
		& Hard          \\ \hline
		Weekend Coverage (WC)        & \makecell[l]{every weekend needs to have
			exactly \\ one clinician that covers it}                                        
		& Hard          \\ \hline
		Min/Max (MM)                  & \makecell[l]{for a given service, each
			clinician can only \\ work between the minimum and maximum \\ number of allowed
			blocks} & Hard          \\ \hline
		No Consecutive Blocks (NCB)   & \makecell[l]{any clinician should not work \\
			two consecutive blocks, across all services}                                   
		& Hard          \\ \hline
		No Consecutive Weekends (NCW) & \makecell[l]{any clinician should not work two
			consecutive \\ weekends}                                                       
		& Hard          \\ \hline
		Equal Weekends (EW)          & \makecell[l]{weekends should be equally
			distributed \\ between clinicians}                                              
		& Hard          \\ \hline
		Equal Holidays (EH)          & \makecell[l]{long weekends should be equally
			distributed \\ between clinicians}                                              
		& Hard          \\ \hline
		Block Requests (BR)         & \makecell[l]{each clinician can request to be
			off service \\ during certain blocks throughout the year}                       
		& Soft          \\ \hline
		Weekend Requests (WR)        & \makecell[l]{each clinician can request to be
			off service \\ during certain weekends throughout the year}                     
		& Soft          \\ \hline
		Block-Weekend Adjacency (BWA)  & \makecell[l]{the block and weekend
			assignments of a given \\ clinician should be adjacent}                         
		& Soft
	\end{tabular}
	\caption{Summary of the constraints for the clinician scheduling problem}\label{tbl:constraint-summary}
\end{table}
